
---
title: "NYPD Shooting Incident Data Analysis"
author: "Sid"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(ggplot2)
```

## Introduction

This report analyzes the NYPD Shooting Incident data. The goal is to explore trends in shooting incidents, including temporal trends, geographical distribution, and perpetrator/victim demographics. We will identify potential biases in the data and their implications on the analysis.

## Data Loading and Preprocessing

```{r load-data}
# Load the data

urlin <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
data <- read.csv(urlin)

# Display the first few rows of the dataset
head(data)
```

## Data Cleaning

```{r clean-data}
# Convert dates and categorize necessary fields
data$OCCUR_DATE <- as.Date(data$OCCUR_DATE, format="%m/%d/%Y")
data$BORO <- as.factor(data$BORO)
data$PERP_AGE_GROUP <- as.factor(data$PERP_AGE_GROUP)
data$PERP_SEX <- as.factor(data$PERP_SEX)
data$PERP_RACE <- as.factor(data$PERP_RACE)
data$VIC_AGE_GROUP <- as.factor(data$VIC_AGE_GROUP)
data$VIC_SEX <- as.factor(data$VIC_SEX)
data$VIC_RACE <- as.factor(data$VIC_RACE)

# Handle missing data: imputation or removal
data <- na.omit(data)  # Removing rows with any NA values for simplicity

# Drop unnecessary columns
data <- select(data, -c(LOC_OF_OCCUR_DESC, LOC_CLASSFCTN_DESC, LOCATION_DESC, Lon_Lat))

# Summary of cleaned data
summary(data)
```

## Data Analysis and Visualization

### Distribution of Incidents by Borough

```{r incidents-by-borough}
ggplot(data, aes(x=BORO)) +
  geom_bar() +
  labs(title="Number of Shooting Incidents by Borough", x="Borough", y="Count")
```

### Trend Over Time

```{r trend-over-time}
data$Year <- year(data$OCCUR_DATE)
ggplot(data, aes(x=Year)) +
  geom_line(stat="count") +
  labs(title="Shooting Incidents Over Time", x="Year", y="Number of Incidents")
```

## Statistical Analysis

### Simple Model: Linear Trend of Incidents Over Time

```{r linear-model}
# Summarize incidents by year
annual_data <- data %>%
  group_by(Year) %>%
  summarise(Incidents = n(), .groups = 'drop')

# Fit linear model to incidents over time
model <- lm(Incidents ~ Year, data = annual_data)

# Summary of the model
summary(model)

# Create a dataframe for predictions
prediction_data <- data.frame(Year = min(annual_data$Year):max(annual_data$Year))
prediction_data$Predicted_Incidents <- predict(model, newdata = prediction_data)

# Plot actual data and model predictions
## Plot actual data and model predictions
ggplot(annual_data, aes(x = Year, y = Incidents)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_line(data = prediction_data, aes(x = Year, y = Predicted_Incidents), color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Shooting Incidents Over Time with Linear Model Prediction",
       x = "Year",
       y = "Number of Incidents") +
  theme_minimal() +
  geom_point() +
  geom_point(data = prediction_data, aes(x = Year, y = Predicted_Incidents), color = "red") +
  scale_x_continuous(breaks = seq(min(annual_data$Year), max(annual_data$Year), by = 1))  # Adjust x-axis breaks if necessary

## Check model diagnostics with a residual plot
ggplot(model, aes(.fitted, .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals")

```

## Conclusion

This report provided an overview of shooting incidents in NYC, explored the trends over time, and performed a basic modeling to understand the trajectory of incidents. Potential biases include underreporting and the limitations of demographic categorization.
